\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-08 papers/esop2010/main.tex
version submitted 15h

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions.  We also define a weakest
  precondition calculus for this language which generates proof obligations in
  higher-order logic.  To our knowledge, this is the first time that effect
  polymorphism is put to use in program verification. Additionally, we propose
  a region-based analysis to exclude aliasing, in order to obtain much simpler
  proof obligations. Several examples demonstrate the practicality of the
  system. A prototype implementation exists and has been used to prove the
  correctness of these and more examples.  \end{abstract}

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} introduces our language using a motivating example.
Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
programming language and the weakest precondition calculus.  In
section~\ref{sec:examples} we present our implementation and some
examples. Finally, section~\ref{sec:related} discusses related work.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays.

@@

To avoid this situation, we additionally introduce a simple means to exclude
programs with aliasing. The instantiation of effect variables is restricted to
the ones which do not create aliasing situations. Our system, with aliasing
exclusion enabled, is inappropriate for reasoning about shared mutable
structures. However, it is interesting to see that this alias exclusion
restriction, in many cases, corresponds to a natural condition imposed on
higher-order functions that operate on some data structure. It is a common
requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration functions,
for example, that the functional argument must not modify the structure to be
iterated on. In our system, this fact follows from the restriction on effect
instantiations, as will be seen in section~\ref{sec:syntax}. In the example,
the array {\tt a} of the iterating function {\tt iter} must not be modified by
its argument.

@@

Finally, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

@@

\subsection{Alias Restriction}

@@

In alias-free programs, the operations $\oplus$ and $|_ε$ on stores can be
executed {\em statically}, i.e. just by looking at the domains of the stores.
Also, we statically know which value (or which named component) the accessor
{\tt !!} will get from a store.

@@

\subsection{Correctness}

@@

\subsection{A simple memoization function}

@@

\subsection{A More Powerful Memoization Function}

@@

  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
    { forall (k : [[int ->{table} int ]]),
        { realizes f0 k} ff k { r : realizes f0 r} }
    let rec f (x : int) =
      { stores f0 !!table }
      if mem x !table then get x !table
      else
        let z = ff f x in
        table := set x z !table; z
      {r : r = f0 x /\ stores f0 !!table}
    in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specification}
  \label{fig:ymemo}
\end{figure}

Our second, more complicated example is the recursive version of {\tt memo},
dubbed {\tt ymemo}. We want to be able to define a recursive function whose
recursive calls are automatically memoized. The basic idea of {\tt ymemo} is
analogous to the fixpoint combinator in λ-calculus. As an example, here is how
one would implement a memoized Fibonacci function
using ymemo (ignoring the logic argument $f0$):
\begin{ocaml}
  let fib = 
    ymemo (fun g x -> if x <= 1 then x else g (x-1) + g (x-2))
\end{ocaml}
The anonymous function uses its first argument {\tt f} to do recursive calls.

{\tt ymemo} achieves this by defining a recursive function {\tt f} inside its
body (see Fig~\ref{fig:ymemo}). This function, when called with an integer
argument {\tt x} first checks if the table contains {\tt x} and if not, calls
{\tt ff} {\em with {\tt f} itself} and x as argument. As before, it stores the
obtained result in the table and returns it. Now, the return value of {\tt
ymemo} is precisely this recursive function {\tt f}. The only difference with
the equivalent \ocaml\ code is the usage of a logical argument {\tt f0} which
stands for the (mathematical) function to be defined.

@@

A prototype implementation of the presented system, called ``\who'', is freely
available~\cite{KanigFilliatre09wml}\footnote{It is available at
\url{http://www.lri.fr/~kanig/who.html}.}. It implements the described
language with type, region and effect polymorphism, higher-order logic and
store types in the logic. It takes as an input a fully annotated program and
outputs proof obligations which have to be proved to guarantee the correctness
of the program with respect to its specification. Currently, only output to
the interactive proof assistant Coq has been implemented. Our logic is a
subset of Coq's, so the translation is particularly simple.  The
implementation consists of less than 3000 lines of \ocaml\ code, including a
limited form of type and effect inference.

@@

Honda, Yoshida and
Berger~\cite{Honda05anobservationally,BergerHondaYoshida05aliasing} present a
logic for imperative higher-order programs. They present their calculus using
rules in Hoare logic style, while we presented a \wpre-calculus. Their system
does not provide the same modularity than ours, as one can not reason
modularly about side effects of functional arguments. It also has the drawback
of being difficult to implement, due to the presence of a large number of
non-structural rules. Our system, starting from a correctly annotated program,
is syntax-directed and thus entirely automated. Their logic of proof
obligations is non-standard, while we obtain obligations in standard higher-order
logic.

They deal with aliasing using equalities and disequalities between
pointers, which we chose not to do.

They can reason about the classic {\em
Landin's knot}, which consists in obtaining recursion by storing a function in
a reference which calls this same reference. Our typing restriction prohibits
this particular kind of programs.  Honda, Yoshida and Berger prove a
completeness result for their system. We have proved no such result.

%, using a slightly more
%expressive language than ours. However, their calculus can
%deal easily with functions in references, even if the stored function
%reads or modifies the reference it is stored in. While, with the
%appropriate extensions, one can reason about functions in references
%in our system, this particular kind of programs cannot be reasoned
%about.  Their extended
%system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
%drawbacks compared to ours, but additionally deals with aliasing
%situations.

@@

We see two possibilities to deal with aliasing beyond plain usage of pointer
disequalities. The first one is to modify the effect system itself to allow
more precise tracking of aliasing. In the setting with the aliasing
restriction, our regions actually become {\em singleton regions}, regions
containing exactly one memory location. Other, more powerful effect systems
allow for {\em group regions}, with the possibility to temporarily {\em focus
} on a particular memory cell in that region. We see no fundamental difficulty
in adopting such an effect system, but haven't worked out the details. A
second possibility is to deal with aliasing via equalities and disequalities
over pointers, using regions to obtain some separation automatically. On
itself, we believe that this technique does not scale, due to the explosion of
needed disequalities.  However, in combination with other separation
techniques, for example the Burstall-Bornat memory model~\cite{bornat00mpc}
which separates automatically objects of different structural types, this
approach has worked well applied to first-order programs.

>>>>

\begin{abstract} We introduce a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions.  We also define a weakest
  precondition calculus for this language which generates proof obligations in
  higher-order logic.  To our knowledge, this is the first time that effect
  polymorphism is put to use in program verification. Additionally, we propose
  a region-based analysis to exclude aliasing, in order to obtain much simpler
  proof obligations. Several examples demonstrate the practicality of the
  system. A prototype implementation exists and has been used to prove the
  correctness of these and more examples.  
\end{abstract}

%TODO overfull boxes

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} introduces our language using a motivating example.
Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
programming language and the weakest precondition calculus.
Section~\ref{sec:alias} shows how two modifications to the type system can
eliminate aliasing. In section~\ref{sec:examples} we present our
implementation and some examples. Finally, section~\ref{sec:related} discusses
related work.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
the logic, and similarily access to the {\tt i}th element is {\tt get i !a} in
the program and {\tt get i !!a} in the logic.

@@

That is why we also propose in this paper a simple technique to exclude
programs with aliasing and to obtain simpler proof obligations. The solution
consists in restricting the instantiation of effect variables. In the example
of Fig.~\ref{fig:iter}, this means that {\tt e} could not be instantiated with
an effect containing {\tt a}.  With aliasing exclusion enabled, our system is
of course inappropriate for reasoning about shared mutable structures.
However, it is interesting to see that this alias exclusion restriction, in
many cases, corresponds to a natural condition imposed on higher-order
functions that operate on some data structure. It is a common
requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration functions,
for example, that the functional argument must not modify the structure to be
iterated on. In our system, this fact follows from the restriction on effect
instantiations, as will be seen in section~\ref{sec:alias}.

It is important to note that neither our specification language nor its
weakest precondition calculus are restricted to alias-free programs. The
aliasing exclusion is only a means to get simpler proof obligations whenever
possible.

@@

The store types $\efft{ε}$ describe a portion of the store and are basically
mappings from region variables to values. The domain of the mapping is the
effect $ε$, where effect variables stand for unknown parts of the domain. To
dea with objects of these types, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

@@

We do not give a correctness proof for this calculus; this proof would be very
similar to and actually simpler than the proof given in the next section.

\section{The Alias Restriction}
\label{sec:alias}

@@

In alias-free programs, the operations $\oplus$ and $|_ε$ on stores can be
executed {\em statically}, i.e. just by looking at the domains of the stores.
Also, we statically know which value (or which named component) the accessor
{\tt !!} will get from a store. Therefore, reasoning about stores and
equalities between regions can be completely eliminated.

@@

\paragraph{Correctness.}

@@

@@

\subsection{Memoization of Recursive Functions}

@@

  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
    { forall (k : [[int ->{table} int ]]),
        { realizes f0 k} ff k { r : realizes f0 r} }
    let rec f (x : int) =
      { stores f0 !!table }
      if mem x !table then get x !table
      else
        let z = ff f x in
        table := set x z !table; z
      {r : r = f0 x /\ stores f0 !!table}
    in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo} function and its specification}
  \label{fig:ymemo}
\end{figure}

Our second, more complicated example is a version of {\tt memo} for recursive
functions, dubbed {\tt ymemo}. We want to be able to define a recursive
function whose recursive calls are automatically memoized. The basic idea of
{\tt ymemo} is analogous to the fixpoint combinator in λ-calculus, hence the
name. As an example, here is how one would implement a memoized Fibonacci
function using {\tt ymemo}:
\begin{ocaml}
  let fib = 
    ymemo (fun g x -> if x <= 1 then x else g (x-1) + g (x-2))
\end{ocaml}
The anonymous function uses its first argument {\tt g} to do recursive calls.

{\tt ymemo} achieves this by defining a recursive function {\tt f} inside its
body (see Fig~\ref{fig:ymemo}). This function, when called with an integer
argument {\tt x} first checks if the table contains {\tt x}, and if not, calls
{\tt ff} {\em with {\tt f} itself} and x as argument. As before, it stores the
obtained result in the table and returns it. Now, the return value of {\tt
ymemo} is precisely this recursive function {\tt f}. The only difference with
the equivalent \ocaml\ code is the usage of a logical argument {\tt f0} which
stands for the (mathematical) function to be defined.

@@

A prototype implementation of the presented system, called ``\who'', is freely
available~\cite{KanigFilliatre09wml}\footnote{It is available at
\url{http://www.lri.fr/~kanig/who.html}.}. It implements the described
language with type, region and effect polymorphism, higher-order logic and
store types in the logic. It is currently restricted to the subset
characterized by the alias restriction. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.  The
implementation consists of less than 3000 lines of \ocaml\ code, including a
limited form of type and effect inference.

@@

Honda, Yoshida and
Berger~\cite{Honda05anobservationally,BergerHondaYoshida05aliasing} present a
logic for imperative higher-order programs. They present their calculus using
rules in Hoare logic style, while we presented a \wpre-calculus. Their system
does not provide the same modularity than ours, as one can not reason
modularly about side effects of functional arguments. It also has the drawback
of being difficult to implement, due to the presence of a large number of
non-structural rules. Our system, starting from a correctly annotated program,
is syntax-directed and thus entirely automated. Their logic of proof
obligations is non-standard, while we obtain obligations in standard
higher-order logic. They deal with aliasing using equalities and disequalities
between pointers.  They can reason about the classic {\em Landin's knot},
which consists in obtaining recursion by storing a function in a reference
which calls this same reference. Our typing restriction prohibits this
particular kind of programs.  Honda, Yoshida and Berger prove a completeness
result for their system. We have proved no such result.

@@

We see two possibilities to deal with aliasing beyond plain usage of pointer
disequalities. The first one is to modify the effect system itself to allow
more precise tracking of aliasing. In the setting with the aliasing
restriction, our regions actually become {\em singleton regions}, regions
containing exactly one memory location. Other, more powerful effect systems
allow for {\em group regions}, with the possibility to temporarily {\em focus
} on a particular memory cell in that region. We see no fundamental difficulty
in adopting such an effect system, but haven't worked out the details. A
second possibility is to use techniques of static analysis to infer disjoint
regions, each possibly containing several pointers~\cite{hubert07hav}.
This approach has proved useful in combination with other separation
techniques, for example the Burstall-Bornat memory model~\cite{bornat00mpc}
which separates automatically objects of different structural types.
